Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-23 doc/asplos13/evaluation.tex

%% UTS: performance comparison
\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.5\textwidth]{figs/uts_compare.pdf}
    \end{center}
    \caption{Performance of in-memory unbalanced tree search. Grappa
        keeps scaling through 128 nodes and beats the XMT with 16
        nodes.}
    \label{fig:uts_compare}
\end{figure}
\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.5\textwidth]{figs/uts_xmt_temp.pdf}
    \end{center}
    \caption{This is XMT uts result, \TODO{overlay the plots}}
    \label{fig:uts_compare_temp}
\end{figure}

@@

\TODO{incorporate UTS chunk size after rerun}
it's important to steal multiple tasks to 
amortize cost of stealing
spread work out quickly in large system

there are times when it's useful to steal lots of stuff, even though
we steal only a few things most of the time.

@@

amortizing cost of stealing
spreading work out more
most of the time we still little  (about 10 avg) but sometimes a lot
(difference with BFS)

@@

We see benefit to limiting the amount of parallelism created by a
recursive loop decoposition. The decomposition threshold parameter the
runtime when to stop creating new tasks and just execute iterations
serially. This allows us to amoritize the overhead of task creation
and potentially exploit locality when data for adjacent iterations is
also adjacent in memory.

@@

% uts threshold
\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.5\textwidth]{figs/uts_threshold.pdf}
    \end{center}
    \caption{Performance of UTS-Mem with varying parallel loop
        threshol, run with 30 nodes, 6 cores per node, 4000 workers,
    \checkme{6M flush ticks}}
    \label{fig:uts_threshold}
\end{figure}

>>>>

%% UTS: performance comparison
\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.5\textwidth]{figs/uts_compare.pdf}
    \end{center}
    \caption{Performance of in-memory unbalanced tree search. Grappa
        keeps scaling through 128 nodes and beats the XMT with 16
        nodes.}
    \label{fig:uts_compare}
\end{figure}

@@

It is important to steal multiple tasks at a time to both amortize the
cost of stealing over the network and to spread out work quickly in a
large system. Figure~\ref{fig:ut_chunksize} shows performance and
stealing statistics for UTS on \checkme{30} nodes as we increase the stealing chunk size. Recall
that a thief will take a number of tasks equal to the minimum of half
the available work or the chunk size; steals fail only when the victim
has fewer than 2 available tasks. As the scheduler is allowed to
steal more work beyond 1 task, we see that performance increases up to 6x. This
shows that the heuristic of stealing the oldest task from victims is
insufficient alone when a tree-structured computation is imbalanced,
as observed in \cite{UTS}. By observing sampled state in the execution
trace, we find that a chunk size as low as 1 allows stealing to
spread the load evenly across the cluster but cores spend much time
underutilized as multiple workers wait for steal replies that
utlimately return little new work.

Performance plateaus before maximum steal amount is limited by the
size of the victim's task queue. This indicates that artificially limiting steals
to \checkme{128} tasks does not limit performance. Although a lower
chunk size limits how quickly work spreads, for sufficient chunk size,
the heuristic of stealing the oldest tasks from victims in tree-based computations allows for
stolen work to expand quickly.

@@

\TODO{(difference with BFS)}

@@

Parallel overhead---in the form of context switches, task spawns, and
synchronization---can reduce the performance benefit of parallelism.
Grappa sees a benefit to limiting the amount of parallelism created by
a recursive loop decomposition. The parallel loop threshold (``parallel
granularity'') parameter tells the runtime when to stop creating new tasks and just execute iterations
sequentially. This allows us to amoritize the overhead of task
creation. In addition, assigning sequential iterations to a single
task provides the potential to exploit locality when data for adjacent iterations is
also adjacent in memory.

@@

% uts threshold
\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.5\textwidth]{figs/uts_threshold.pdf}
    \end{center}
    \caption{Performance of UTS-Mem with varying parallel loop
        threshold, run with 30 nodes, 6 cores per node, 4000 workers,
    \checkme{6M flush ticks}}
    \label{fig:uts_threshold}
\end{figure}

### 2012-07-23 doc/asplos13/intro.tex

Our goals is challenging for two key reasons:

@@

Luckily, many of the important irregular applications (e.g., graph
processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency
to tolerate the latency of data movement. The fully custom Tera
MTA-2~\cite{mta-2} system is a classic example of supporting irregular
applications by using concurrency to hide latencies. It had a large
distributed shared memory with no caches.  On every clock cycle, each
processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory
access latency.  The network was designed with a single word injection
rate that matched the processor clock frequency and sufficient bandwidth
to sustain a reference from every processor on every clock cycle.
Unfortunately, while an excellent match to extremely irregular
applications, the MTA was not cost-effective on applications that could
exploit locality and had very poor single-thread performance, making
it a commercial failure. The Cray XMT approximates the Tera MTA-2,
substantially reducing its cost but not overcoming its narrow range of
applicability.

@@

In this paper we introduce Grappa, a software runtime system that allows
a commodity x86 distributed-memory HPC cluster to be programmed as if it
were a single large shared-memory machine and provides scalable
performance for irregular applications. Grappa is designed to smooth
over some of the performance discontinuities in commodity hardware,
giving good performance when there is little locality to be exploited
while allowing the programmer to exploit it when it is available. Grappa
leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system. An off the
shelf user-mode infiniband device driver stack~\cite{Melonox?} is used
for network interfacing. MPI~\cite{mpi} is used for process setup and
tear down. GASnet~\cite{gasnet} is used as the underlying mechanism for
remote memory reads and writes using active message invocations. To this
commodity hardware and software mix Grappa adds three main software
components: (1) a \emph{lightweight tasking} layer that supports a
context switch in as few as 38ns and distributed load balancing; (2) a
\emph{distributed shared memory} layer that supports customary
operations such as \emph{read and write as well as remote operations}
such as fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message
aggregation} layer that combines short messages to mitigate the
aforementioned problem that commodity networks are designed to achieve
peak bandwidth only on large packet sizes, yet irregular applications
tend to fetch only a handful of bytes at a time.

>>>>

Our goal is challenging for two key reasons:

@@

Luckily, many of the important irregular applications (e.g., graph
processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency
to tolerate the latency of data movement. The fully custom Tera
MTA-2~\cite{tera} \TODO{this is mta-1 citation} system is a classic example of supporting irregular
applications by using concurrency to hide latencies. It had a large
distributed shared memory with no caches.  On every clock cycle, each
processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory
access latency.  The network was designed with a single word injection
rate that matched the processor clock frequency and sufficient bandwidth
to sustain a reference from every processor on every clock cycle.
Unfortunately, while an excellent match to extremely irregular
applications, the MTA was not cost-effective on applications that could
exploit locality and had very poor single-thread performance, making
it a commercial failure. The Cray XMT approximates the Tera MTA-2,
substantially reducing its cost but not overcoming its narrow range of
applicability.

@@

In this paper we introduce Grappa, a software runtime system that allows
a commodity x86 distributed-memory HPC cluster to be programmed as if it
were a single large shared-memory machine and provides scalable
performance for irregular applications. Grappa is designed to smooth
over some of the performance discontinuities in commodity hardware,
giving good performance when there is little locality to be exploited
while allowing the programmer to exploit it when it is available. Grappa
leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system. An off the
shelf user-mode infiniband device driver stack~\cite{Melonox?} is used
for network interfacing. MPI is used for process setup and
tear down. GASnet~\cite{gasnet:spec} is used as the underlying mechanism for
remote memory reads and writes using active message invocations. To this
commodity hardware and software mix Grappa adds three main software
components: (1) a \emph{lightweight tasking} layer that supports a
context switch in as few as 38ns and distributed load balancing; (2) a
\emph{distributed shared memory} layer that supports customary
operations such as \emph{read and write as well as remote operations}
such as fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message
aggregation} layer that combines short messages to mitigate the
aforementioned problem that commodity networks are designed to achieve
peak bandwidth only on large packet sizes, yet irregular applications
tend to fetch only a handful of bytes at a time.

### 2012-07-23 doc/asplos13/relatedwork.tex

\paragraph{Global memory} Grappa includes a custom implementation of a
software distributed shared memory (DSM) system. Many traditional software DSM
systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that
they are built in software from applications by exploiting the processor's
paging mechanisms, therefore relying heavily on locality. Instead, Grappa,
like other partitioned global address space (PGAS) models, implements its DSM
at the language, rather than system level. Languages such as Chapel
\cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared
structures look like normal memory references. As we describe later, Grappa
chooses a middle ground, where global addresses are explicit in the API and
local accesses are emitted conventionally by the compiler. Similar to DSM
systems, Grappa provides a consistency model that ensure high performance and
with good programmability properties. % Caching is Cache operations are explicit in the  exposed to in the
% API, however, allowing software to optimize their usage.

@@

\paragraph{Multithreading}
Grappa uses multithreading to tolerate memory latency. This is a well known
technique. Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT
\cite{XMT}, Simultaneous multithreading \cite{SMT}, MIT Alewife
\cite{Alewife}, Cyclops \cite{Cyclops}, and even GPUs \cite{fatahalian}. As we
describe in this paper, Grappa use a lightweight user-mode task scheduler to
multiplex thousands of tasks on a single processing core. The large number of
tasks is required because of the extremely large internode latency.

>>>>

\paragraph{Global memory} Grappa includes a custom implementation of a
software distributed shared memory (DSM) system. Many traditional software DSM
systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that
they are built in software from applications by exploiting the processor's
paging mechanisms, therefore relying heavily on locality. Instead, Grappa,
like other partitioned global address space (PGAS) models, implements its DSM
at the language, rather than system level. Languages such as Chapel
\cite{Chamberlain:2007}, X10 \cite{X10:2005}, and UPC \cite{upc:2005} make accesses to shared
structures look like normal memory references. As we describe later, Grappa
chooses a middle ground, where global addresses are explicit in the API and
local accesses are emitted conventionally by the compiler. Similar to DSM
systems, Grappa provides a consistency model that ensure high performance and
with good programmability properties. % Caching is Cache operations are explicit in the  exposed to in the
% API, however, allowing software to optimize their usage.

@@

\paragraph{Multithreading}
Grappa uses multithreading to tolerate memory latency. This is a well known
technique. Hardware implementations include the Tera MTA \cite{tera:mta1}, Cray XMT
\cite{feo:xmt}, Simultaneous multithreading \cite{tullsen:smt}, MIT Alewife
\cite{agarwal:alewife}, Cyclops \cite{almasi:cyclops}, and even GPUs \cite{gpus}. As we
describe in this paper, Grappa use a lightweight user-mode task scheduler to
multiplex thousands of tasks on a single processing core. The large number of
tasks is required because of the extremely large internode latency.
